DNA Methylation Analyses

### 2023-02-27 DNA-methylation.Rmd
meth - reduced image size
Update DNA-methylation.Rmd
clarified descriptions, added links to bismark manual where appropriate

The purpose of this tutorial is to provide an example of taking Bisulfite treated DNA sequence data through identification of methylation at the loci and gene level. The principle behind creating bisulfite treated DNA libraries is that when cytosines are not methylated they will be converted to uracil, which ultimately result in thymine. When aligning to a reference genome the percent methylation at a given CpG loci is determined by examine the ratio to thymines and cytosines. For example if there is a given CpG loci with 30% of the alignment with the cytosine on the reference being and thymine, one would consider this cytosine loci (CpG) to be 70% methylated.

@@

![bp-chart](images/meth-bp-chart.png)

@@

You should expect to prepared genome with new directories indluding

@@

For doing the alignment of the trim reads to the genome the following code structured is used

@@

Files are then deduplicated if they are whole genome bisulfite treated samples. This command will deduplicate the Bismark alignment BAM file and remove all reads but one which align to the the very same position and in the same orientation. This step is recommended for whole-genome bisulfite samples, but should not be used for reduced representation libraries such as RRBS, amplicon or target enrichment libraries.

@@

This will create a file with the suffix `deduplicated.bismark.cov.gz` (example shown below in uncompressed format). And while a bedGraph file is generated, in this workflow a bedGraph will be generated later that uses information from merged strands (as opposed to this output).

@@

A genome-wide cytosine report is generated, with this code as an example:

@@

Generating a file `.CpG_report.merged_CpG_evidence.cov`

@@

From here it could be useful to do some simple file reorganizaion to obtain bedgraph or tab files for downstream analysis (eg bedtools, GLMs).

    #creating bedgraphs post merge
    for f in *merged_CpG_evidence.cov
    do
      STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
      cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4}}' \
      > "${STEM}"_10x.bedgraph
    done

@@

    #creating tab files with raw count for glms
    for f in *merged_CpG_evidence.cov
    do
      STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
      cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4, $5, $6}}' \
      > "${STEM}"_10x.tab
    done

>>>>

The purpose of this tutorial is to provide an example of identifying loci- and gene-level methylation in bisulfite-treated DNA sequence data. The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data. When aligning to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines. For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated.

@@

```{r meth-bp, echo = FALSE, out.width = "70%", fig.align = "center"}
knitr::include_graphics("images/meth-bp-chart.png")
```

For WGBS and MBDBS data, it is general practice to hard-trim 10 bp from the beginning and end of each read. [RRBS data often needs different trimming specifications](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html#ix-notes-about-different-library-types-and-commercial-kits).

@@

You should expect to prepared genome with new directories including

@@

For aligning the trimmed reads to the genome the following code structure is used

@@

Files are then deduplicated if they are whole genome bisulfite treated samples. This command will deduplicate the Bismark alignment BAM file and remove all reads that align to the the very same position and in the same orientation except for one. This step is recommended for whole-genome bisulfite samples, but **should not** be used for reduced representation libraries such as RRBS, amplicon, or target enrichment libraries. The [`bismark` manual](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html#ix-notes-about-different-library-types-and-commercial-kits) provides additional information.

@@

This will create a file with the suffix `deduplicated.bismark.cov.gz` (example shown below in uncompressed format). The resultant bedGraph file contains information from *unmerged* strands, meaning reads aligning to forward and reverse strands in the same position are not combined. Later in this workflow, a bedGraph with *merged* information will be created.

@@

A genome-wide cytosine report is generated (including merged bedGraph file), with this code as an example:

@@

Generating a file `.CpG_report.merged_CpG_evidence.cov` (merged bedGraph file)

@@

From here it could be useful to do some simple file reorganizaion to obtain bedGraph or tab files for downstream analysis (eg bedtools, GLMs).

Creating bedGraphs post-merge at a specific coverage threshold (ex. 10x or 5x):

    for f in *merged_CpG_evidence.cov
    do
      STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
      cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4}}' \
      > "${STEM}"_10x.bedgraph
    done

@@

Creating tab files with raw count data for GLMs at 10x and 5x coverage:

    for f in *merged_CpG_evidence.cov
    do
      STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
      cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4, $5, $6}}' \
      > "${STEM}"_10x.tab
    done
